\section{Simulation}
\label{sec:simul}

The model performance is averaged on $300$ synthetic datasets with individuals $n=100$ of each. We design a similar model as in Sec \ref{sec:applc}. The true parameters are $\boldsymbol{\beta}=(2,1,-0.3,1.5)^T$, $\boldsymbol{\alpha}=(-1.65,0.15,1.8,-0.05,0.4)^T$, $\boldsymbol{\gamma}=(-0.75,-1.5,-2)^T$, the variances of errors in the linear mixed effects model of $Y$ is $\sigma=0.75$, and the covariance between two vectors of random effects as $0.3$ for all individuals. Choose $\text{LLOQ}=2$, let the secondary parameters $d_1=0.5$, $d_2=0.15$. In the Cox PH models, let the baseline hazards models to be $h_0=800$. The simulation of Cox and Weibull regression models are in Table \ref{tbl:siml}.
\begin{table}[ht]
\centering
\caption{Simulation results of Cox and Weibull regression models using h-likelihood}
\begin{tabular}{rr|rrr|rrr}
  \hline
   & & \multicolumn{3}{c|}{Cox} &  \multicolumn{3}{c}{Weibull} \\
  \hline
   & True para & Estimates & Std\_Errors & P-values & Estimates & Std\_Errors & P-values \\ 
  \hline
  $\beta_0$ & 2.000 & 2.144 & 0.096  & 0.000 & 2.153 & 0.096  & 0.000 \\ 
  $\beta_1$ & 1.000 & 0.997 & 0.234 & 0.000 & 0.992 & 0.233 & 0.000 \\ 
  $\beta_2$ & -0.300 & -0.311 & 0.159  & 0.050 & -0.309 & 0.158  & 0.051 \\ 
  $\beta_3$ & 1.500 & 1.395 & 0.081  & 0.000 & 1.389 & 0.081  & 0.000 \\ 
   \hline
  $\alpha_0$ & -1.650 & -1.191 & 0.220  & 0.000 & -1.273 & 0.223  & 0.000 \\ 
  $\alpha_1$ & 0.150 & \textcolor{red}{2.579} & 0.079  & 0.000 & \textcolor{red}{2.751} & 0.076  & 0.000 \\ 
  $\alpha_2$ & 1.800 & 1.522 & 0.376 & 0.000 & 1.554 & 0.382 & 0.000 \\ 
  $\alpha_3$ & -0.050 & -0.071 & 0.038  & \textcolor{red}{0.064} & -0.066 & 0.039  & \textcolor{red}{0.089} \\ 
  \hline
  $\gamma_0$ & -0.750 & -0.644 & 0.132  & 0.001 & -0.463 & 0.139  & 0.001 \\ 
  $\gamma_1$ & -1.500 & \textcolor{red}{-0.614} & 0.134  & 0.015 & \textcolor{red}{-0.310} & 0.240  & \textcolor{red}{0.196} \\ 
  $\gamma_2$ & -2.000 & \textcolor{red}{-1.561} & 0.162 & 0.004 & \textcolor{red}{-1.008} & 0.135  & 0.000 \\ 
   \hline
\end{tabular}
\label{tbl:siml}
\end{table}


The simulation study shows that the parametric Weibull regression models are not more efficient than semi-parametric Cox models as the standard errors of parameters are very similar in both models. There is no obviously higher accuracy in either joint model. Most of the estimates are close to the true parameters; however, three of them ($\alpha_1$, $\gamma_1$, and $\gamma_2$) are not. 

There can be several reasons. Starting values may be far from the optimal which results in an early convergence of the algorithm. As Cox and Weibull are both proportional hazards models, they both have proportional hazards assumption.
Measure the performance of an alternative AFT survival model on the same data, and compare if the alternative model performs better in the three parameters ($\alpha_1$, $\gamma_1$, and $\gamma_2$). If so, the proportional hazards assumption is not valid in the synthetic data.



